---
title: |
  <div class="tight-title">
    <div>ECOM30004/90004</div>
    <div>Time Series Analysis and Forecasting</div>
     <div class="spacer"></div> <!-- blank line -->
    <div class="bold">Tutorial 10</div>

    <div class="spacer"></div> <!-- blank line -->

    <div class="author">Tutor: Pengyun (Kevin) Chen</div>
    <div class="author">University of Melbourne</div>
    <div class="author">Semester 2, 2026</div>
  </div>

  
format:
  revealjs:
    theme: simple
    self-contained: true
    chalkboard: false
    width: 1300
    height: 950
    footer: 'Access Slides: `pengyunc.github.io/TSAF_Week10`'
    title-slide-attributes:
      data-background-color: "#000F46"
    slide-number: true
    center-title-slide: true
    toc: false
    css: styles.css
    callout-appearance: simple
    background-transition: fade
    callout-icon: true
    transition: slide
    scrollable: true
    code-block-height: 800px
    code-line-numbers: true
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js"
    html-math-method: mathjax
    execute:
      echo: true
      eval: true
      enabled: true
---

## Notice

::: {.callout-warning title="Conditions of Use"}

- This document is intended for personal use and is NOT meant to be shared on the internet (e.g. Studocu, ThinkSwap, StudentVIP, Coursehero, etc.)
- Do not share this document with anyone, even if they are enrolled in ECOM30004/90004.
- Official course materials, such as lecture slides, tutorial sheets and prescribed readings should always be your primary point of reference.

:::
<div style="height: 0.2em;"></div>
 - Slides accessible here: `pengyunc.github.io/TSAF_Week10`
 - Download HTML versions here: [Dropbox Link](https://www.dropbox.com/scl/fo/aoxoud5ex0j42hsrod3dg/ADblAt_jJ9hpESZME9FDcWc?rlkey=0s0oyf1y3uokzvlhbzxe2yy3g&st=kx2yupzr&dl=0)



## This tutorial: Prediction Interval Estimation {auto-animate="true"}
<blue>In this tutorial we use simulation to understand and explore the methods for prediction interval calculation.</blue>
<div style="height: 0.5em;"></div>
:::{.callout-note title="Activities"}
- <blue>Exercises 1 & 2:</blue> Simulating 95% prediction intervals — coverage and average length (width)  
- <blue>Exercise 3:</blue> Investigating the effect of model misspecification on prediction interval properties  
- <blue>Exercise 4:</blue> Simulating prediction intervals with non-normal disturbances  
:::


## Recap: Prediction Intervals {auto-animate="true"}

:::{.fragment}
- Consider an AR(1) model and assume $Y_{n+1}$ has conditional distribution:
$$
Y_{n+1} = \phi_{1}Y_{n} + U_{n+1} 
\quad \text{and} \quad 
Y_{n+1}\mid \mathscr{Y}_{t} \sim \mathrm{N}(\phi_{1}Y_{n}, \sigma_{n}^{2})
$$
$$
\color{blue}{\implies\quad \sigma_{n}^{2}=\operatorname{var}(Y_{n+1}\mid\mathscr{Y}_{n})
   = \operatorname{var}(U_{n+1}\mid\mathscr{Y}_{n})}
$$
where $U_{t+1}$ is the 1-step-ahead prediction error.
:::

:::{.fragment}
- The <blue>95% confidence interval for $(Y_{n+1}\mid\mathscr{Y}_{n})$</blue> is:
$$\phi_{1}Y_{n}\pm 1.96\times \sigma_{n}$$
:::

:::{.fragment}
- It follows that the <blue>95\% prediction interval for $Y_{n+1}$</blue> is:
$$\quad\quad\color{purple}{\hat{\phi}_{1}}Y_{t}\color{black}{\,\pm\, 1.96\times\,}\color{teal}{\hat{\sigma}_{n}}\color{black}\quad\text{where}\quad \begin{aligned}\color{purple}{\hat{\phi}_{1}}&\color{purple}{\text{ is the usual parameter estimate}}\\
\color{teal}{\hat{\sigma}^{2}_{n}}&\color{teal}{=\frac{1}{n}\sum^{n}_{t=2}\hat{U}_{t}^{2}}\end{aligned}$$
assuming $\operatorname{var}(U_{t}\mid\mathscr{Y}_{t-1})$ is constant for all $t$.
:::


## Recap: Prediction Intervals {auto-animate="true"}
- It follows that the <blue>95\% prediction interval for $Y_{n+1}$</blue> is:
$$\quad\quad\color{purple}{\hat{\phi}_{1}}Y_{n}\color{black}{\,\pm\, 1.96\times\,}\color{teal}{\hat{\sigma}_{n}}\color{black}\quad\text{where}\quad \begin{aligned}\color{purple}{\hat{\phi}_{1}}&\color{purple}{\text{ is the usual parameter estimate}}\\
\color{teal}{\hat{\sigma}^{2}_{n}}&\color{teal}{=\frac{1}{n}\sum^{n}_{t=2}\hat{U}_{t}^{2}}\end{aligned}$$
assuming $\operatorname{var}(U_{t}\mid\mathscr{Y}_{t-1})$ is constant for all $t$.

:::{.fragment}
- This prediction interval contains $Y_{n+1}$ with approximately 95% probability in repeated samples. <teal>$\to$ we will illustrate this with a simulation</teal>
:::
:::{.fragment}
- <blue>This probability is understood in the frequentist sense:</blue>
  - given repeated drawings from the “population” of the data, the proportion of the constructed intervals will include actual $Y_t$ is 95%.
:::

:::{.fragment}
- <blue>Precision:</blue> A narrower prediction interval indicates a more precise forecast.
:::

## Exercises 1 and 2 {auto-animate="true"}
:::{.callout-question}
Simulate the 95% prediction intervals for $Y_{t+h}$ for $h=1,\dots,12$ for the AR process:
$$Y_{t}=0.8Y_{t-1}+U_{t}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,1)$$
Include in the simulation a calculation for the average length of the PIs for each $h$.
:::
<div style="height: 0.3em;"></div>

:::{.callout-tip title="Code Outline"}
:::{.incremental}
- <blue>Repeat the following experiment `reps=1000` times:</blue>
  - Generate $200+12$ draws of $U_t$ from $\mathrm{N}(0,1)$
  - Simulate $200+12$ observations from the AR(1) process
  - Withhold the final 12 observations as test set
  - Estimate the AR(1) model with the training set
  - Compute and store the upper and lower bounds of 95% PI for $h=1,2,\dots,12$
  - Check whether the true values are contained within the 95% PIs 
- <blue>**Exercise 1:** Compute the probability at which the 95% PIs contains the true $Y_{t+h}$ for $h=1,2,\dots,12$ </blue>
- <blue>**Exercise 2:** Compute the average lengths of the 95% PIs for each $h=1,2,\dots,12$ </blue>
:::
:::


## Exercises 1 and 2 {auto-animate="true"}
:::{.callout-question}
Simulate the 95% prediction intervals for $Y_{t+h}$ for $h=1,\dots,12$ for the AR process:
$$Y_{t}=0.8Y_{t-1}+U_{t}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,1)$$
Include in the simulation a calculation for the average length of the PIs for each $h$.
:::
<div style="height: 0.5em;"></div>

:::{.panel-tabset}

## Initialisation
```{r}
library(forecast)

set.seed(30004)       # For reproducibility

reps <- 1000
n <- 200              # Estimation sample size per experiment
hmax <- 12            # Forecast horizon
ar <- 0.8             # AR(1) true coefficient

# Storage for inclusion indicators
Included <- matrix(nrow=reps, ncol=hmax)

# Storage for length of each interval
Length <- matrix(nrow=reps, ncol=hmax)    
```

## Simulation
```{r}
for (r in 1:reps){
  U <- rnorm(n+hmax)          # N(0,1) prediction errors  
  Y <- arima.sim(n=n+hmax, model=list(ar=ar), innov=U)  # Simulate data
  Yf <- Y[(n+1):(n+hmax)]     # Forecast sample
  Y <- Y[1:n]                 # Estimation sample

  # Estimate AR(1) model and generate forecasts and PIs
  eq <- Arima(Y, order=c(1,0,0))
  eqf <- forecast(eq, h=hmax)  
  
  Lower <- eqf$lower[,"95%"]       # lower bound of 95% PIs
  Upper <- eqf$upper[,"95%"]       # upper bound of 95% PIs

  Included[r,] <- 1*(Lower<Yf & Upper>Yf)     # Check inclusion
  Length[r,] <- Upper-Lower                   # Lengths of PIs
}
```

## Extracting PIs Explained
```{r}
eqf <- forecast(eq, h=hmax)
print(eqf)
```
<div style="height: 0.5em;"></div>
```{r}
# Extract the lower bounds for 95% PI for h=1,2,...,12
print(eqf$lower)  # eqf$lower[,"95%"]
```

## Results
```{r}
# Record 1 if true value contained within 95% PI, 0 otherwise
print(head(Included))   
# Recorded as difference between upper and lower bounds of 95% PI
print(head(Length))     
```
:::


## Exercises 1 and 2 {auto-animate="true"}
:::{.callout-question}
Simulate the 95% prediction intervals for $Y_{t+h}$ for $h=1,\dots,12$ for the AR process:
$$Y_{t}=0.8Y_{t-1}+U_{t}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,1)$$
Include in the simulation a calculation for the average length of the PIs for each $h$.
:::
<div style="height: 0.5em;"></div>

:::{.panel-tabset}

## Exercise 1 - Coverage

:::{.callout-question}
Compute the probabilities at which the 95% PIs contains the true $Y_{t+h}$ for $h=1,\dots,12$.
:::
<div style="height: 0.5em;"></div>
```{r}
Coverage <- apply(Included, 2, mean) * 100
names(Coverage) <- paste0("h=", 1:hmax)
print(Coverage)
```

<div style="height: 0.5em;"></div>
<blue>The simulated coverages are slightly below 95%?</blue>

- arises partly because of simulation error (only 1000 simulations)
- also partly because the forecast intervals are based on model estimates from $n=200$ sample sizes which do not yield the exact model coefficients.

## Exercise 2 - Lengths
:::{.callout-question}
Compute the average lengths of the 95% PIs for each $h=1,2,\dots,12$.
:::
<div style="height: 0.5em;"></div>
```{r}
AvgLength <- round(apply(Length, 2, mean), 2)
names(AvgLength) <- paste0("h=", 1:hmax)
print(AvgLength)
```
<div style="height: 0.4em;"></div>
- The average length of the interval is increasing in $h$, which reflects the formula we have seen in lectures for the variances for $h$-step-ahead forecasts.
$$\textstyle\operatorname{var}(U_{t\mid t-h})=\sum^{h-1}_{j=0}\phi_{1}^{2j}\sigma^{2}\qquad\color{teal}{\boxed{\leftarrow\;\textsf{\small{}From Lecture 9}\;}}$$
- <blue>Intuition: </blue>it is *more difficult* to forecast further into the future.

:::


## Exercise 3 {auto-animate="true"}
:::{.callout-question}
Now change the simulation to demonstrate the effects of model misspecification on the properties of the prediction intervals. $\to$ **Simulate AR(1) data, but forecast with MA(1)**

<blue>What are the implications for the prediction intervals? Explain the results for the lengths of the prediction intervals for $h>1$.</blue>
:::

:::{.panel-tabset}
## Initialisation Code
Code for initialisation is the same as in Exercises 1 and 2.
```{r}
set.seed(30004)       # For reproducibility

reps <- 1000
n <- 200              # Estimation sample size per experiment
hmax <- 12            # Forecast horizon
ar <- 0.8             # AR(1) true coefficient

# Storage for inclusion indicators
Included <- matrix(nrow=reps, ncol=hmax)

# Storage for length of each interval
Length <- matrix(nrow=reps, ncol=hmax)   
```
## Estimation Code
The only change is to the estimation command, replacing AR(1) with MA(1).
```{r}
for (r in 1:reps){
  U <- rnorm(n+hmax)          # N(0,1) prediction errors  
  Y <- arima.sim(n=n+hmax, model=list(ar=ar), innov=U)  # Simulate data
  Yf <- Y[(n+1):(n+hmax)]     # Forecast sample
  Y <- Y[1:n]                 # Estimation sample

  # Estimate MA(1) model and generate forecasts and PIs
  eq <- Arima(Y, order=c(0,0,1))   ### <- MISSPECIFIED!!!
  eqf <- forecast(eq, h=hmax)  
  
  Lower <- eqf$lower[,"95%"]       # lower bound of 95% PIs
  Upper <- eqf$upper[,"95%"]       # upper bound of 95% PIs

  Included[r,] <- 1*(Lower<Yf & Upper>Yf)     # Check inclusion
  Length[r,] <- Upper-Lower                   # Lengths of PIs
}
```
<div style="height: 0.5em;"></div>

## Results: Coverage
```{r}
# Coverage: proportion of inclusion at each h 
Coverage <- apply(Included, 2, mean)*100
names(Coverage) <- paste0("h=",1:hmax)
print(Coverage)
```
<div class="horizontal-line"></div>

<blue>When the model is misspecified:</blue>

- the coverage rates of the prediction intervals decrease below 90% as $h$ increases. 
- the true value is included within the interval less frequently than the desired 90%.


## Results: Lengths
```{r}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```

<div class="horizontal-line"></div>
The average lengths of the 95% PIs are the same for $h=2,\dots,12$.

:::

## Exercise 3 {auto-animate="true"}
<blue>**What's going on?**</blue>
```{r echo=FALSE}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```
<div class="horizontal-line"></div>

:::{.fragment}
:::{.callout-tip}
<green>Consider the MA(1) model:</green> for any $t$ we have
$$Y_{t}=U_{t}+\theta_{1}U_{t-1}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,\sigma^{2})$$
To compute the 95% PI for $Y_{t+h}$, we need to know the distribution for $Y_{t+h}\mid\mathscr{Y}_{t}.$
:::
:::
<div style="height: 0.3em;"></div>
:::{.fragment}
<blue>When $h=1$:</blue>
$$Y_{t+1}=U_{t+1}+\theta_{1}U_{t}$$
:::


:::{.fragment}
$$\implies\quad \mathbb{E}(Y_{t+1}\mid\mathscr{Y}_{t})=\underbrace{\mathbb{E}(U_{t+1}\mid\mathscr{Y}_{t})}_{=\mathbb{E}(U_{t+1})=0}+\theta_{1}\underbrace{\mathbb{E}(U_{t}\mid\mathscr{Y}_{t})}_{=u_t}$$
:::

:::{.fragment}
$$\implies\quad \mathbb{E}(Y_{t+1}\mid\mathscr{Y}_{t})=\theta_{1}u_{t}$$
where $u_{t}$ is the realised value of $U_t$ (it is a constant).
:::


## Exercise 3 {auto-animate="true"}
<blue>**What's going on?**</blue>
```{r echo=FALSE}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```
<div class="horizontal-line"></div>

:::{.callout-tip}
<green>Consider the MA(1) model:</green> for any $t$ we have
$$Y_{t}=U_{t}+\theta_{1}U_{t-1}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,\sigma^{2})$$
To compute the 95% PI for $Y_{t+h}$, we need to know the distribution for $Y_{t+h}\mid\mathscr{Y}_{t}.$
:::
<div style="height: 0.3em;"></div>


<blue>When $h=1$:</blue>

<blue>Mean of $Y_{t+1}\mid\mathscr{Y}_{t}$:</blue> $\mathbb{E}(Y_{t+1}\mid\mathscr{Y}_{t})=\theta_{1}u_{t}$.

:::{.fragment}
<div class="horizontal-line"></div>
<purple>How about the variance?</purple>
:::

:::{.center}

[$\displaystyle\operatorname{var}(Y_{t+1}\mid\mathscr{Y}_{t})=\underbrace{\operatorname{var}(U_{t+1}\mid\mathscr{Y}_{t})}_{=\operatorname{var}(U_{t+1})=\sigma^{2}}+\underbrace{\;\;\operatorname{var}(\theta_{1}U_{t}\mid\mathscr{Y}_{t})\;\;}_{=0,\; \theta_{1}u_{t}\text{ is a constant}}$]{.fragment}[$=\sigma^{2}$]{.fragment}
:::


## Exercise 3 {auto-animate="true"}
<blue>**What's going on?**</blue>
```{r echo=FALSE}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```
<div class="horizontal-line"></div>

:::{.callout-tip}
<green>Consider the MA(1) model:</green> for any $t$ we have
$$Y_{t}=U_{t}+\theta_{1}U_{t-1}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,\sigma^{2})$$
To compute the 95% PI for $Y_{t+h}$, we need to know the distribution for $Y_{t+h}\mid\mathscr{Y}_{t}.$
:::
<div style="height: 0.3em;"></div>


<blue>When $h=1$:</blue>

<blue>Mean of $Y_{t+1}\mid\mathscr{Y}_{t}$:</blue> $\mathbb{E}(Y_{t+1}\mid\mathscr{Y}_{t})=\theta_{1}u_{t}$.

<purple>Variance of $Y_{t+1}\mid\mathscr{Y}_{t}$:</purple> 
$\operatorname{var}(Y_{t+1}\mid\mathscr{Y}_{t})=\sigma^{2}$

:::{.fragment}
<div class="horizontal-line"></div>

Given that $Y_{t+1}\mid\mathscr{Y}_{t}=U_{t+1}+\theta_{1}u_{t}$ and $U_{t+1}\sim\mathrm{N}(0,\sigma^{2})$,
:::

:::{.center}
[$\implies Y_{t+1}\mid\mathscr{Y}_{t}\sim\mathrm{N}(\theta_{1}u_{t},\sigma^{2})\quad$]{.fragment} [$\color{teal}{\implies \textsf{the 95\% PI is}\quad \hat{\theta}_{1}u_{t}\pm 1.96\hat{\sigma}}$]{.fragment}
:::


## Exercise 3 {auto-animate="true"}
<blue>**What's going on?**</blue>
```{r echo=FALSE}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```
<div class="horizontal-line"></div>

:::{.callout-tip}
<green>Consider the MA(1) model:</green> for any $t$ we have
$$Y_{t}=U_{t}+\theta_{1}U_{t-1}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,\sigma^{2})$$
To compute the 95% PI for $Y_{t+h}$, we need to know the distribution for $Y_{t+h}\mid\mathscr{Y}_{t}.$
:::
<div style="height: 0.2em;"></div>

<blue>When $h=1$, the 95% prediction interval is $\hat{\theta}_{1}u_{t}\pm 1.96\hat{\sigma}$.</blue>

[<blue>When $h=2$,</blue>]{.fragment} [we have $Y_{t+2}\mid\mathscr{Y}_{t}=U_{t+2}+\theta_{1}U_{t-1}$.]{.fragment}

:::{.incremental}
- **Mean:** $\;\;\displaystyle\mathbb{E}(Y_{t+2}\mid\mathscr{Y}_{t})=\underbrace{\mathbb{E}(U_{t+2}\mid\mathscr{Y}_{t})}_{=\mathbb{E}(U_{t+2})=0}+\theta_{1}\underbrace{\mathbb{E}(U_{t+1}\mid\mathscr{Y}_{t})}_{=\mathbb{E}(U_{t+1})=0}=0$

- **Variance:** $\;\;\displaystyle\mathrm{var}(Y_{t+2}\mid\mathscr{Y}_{t})=\underbrace{\mathrm{var}(U_{t+2}\mid\mathscr{Y}_{t})}_{=\operatorname{var}(U_{t+2})=\sigma^{2}}+\underbrace{\mathrm{var}(\theta_{1}U_{t+1}\mid\mathscr{Y}_{t})}_{=\theta_{1}^{2}\operatorname{var}(U_{t+1})=\theta_{1}^{2}\sigma^{2}}=(1+\theta_{1}^{2})\sigma^{2}$

- **95% PI:** $\;\;$Since $Y_{t+2}\mid\mathscr{Y}_{t}\sim\mathrm{N}(0,(1+\theta_{1}^{2})\sigma^{2})\;\;\implies\;\; 0\pm1.96\times \sqrt{(1+\hat{\theta}_{1}^{2})\hat{\sigma}^{2}}$
<div style="height: 1em;"></div>
:::


## Exercise 3 {auto-animate="true"}
<blue>**What's going on?**</blue>
```{r echo=FALSE}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```
<div class="horizontal-line"></div>

:::{.callout-tip}
<green>Consider the MA(1) model:</green> for any $t$ we have
$$Y_{t}=U_{t}+\theta_{1}U_{t-1}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,\sigma^{2})$$
To compute the 95% PI for $Y_{t+h}$, we need to know the distribution for $Y_{t+h}\mid\mathscr{Y}_{t}.$
:::
<div style="height: 0.2em;"></div>

<blue>When $h=1$, the 95% prediction interval is $\hat{\theta}_{1}u_{t}\pm 1.96\hat{\sigma}$.</blue>

<blue>When $h\ge 2$,</blue> we have $Y_{t+h}\mid\mathscr{Y}_{t}=U_{t+h}+\theta_{1}U_{t+h-1}$.

- **Mean:** $\;\;\displaystyle\mathbb{E}(Y_{t+h}\mid\mathscr{Y}_{t})=\underbrace{\mathbb{E}(U_{t+h}\mid\mathscr{Y}_{t})}_{=\mathbb{E}(U_{t+h})=0}+\theta_{1}\underbrace{\mathbb{E}(U_{t+h-1}\mid\mathscr{Y}_{t})}_{=\mathbb{E}(U_{t+h-1})=0}=0$

- **Variance:** $\;\;\displaystyle\mathrm{var}(Y_{t+h}\mid\mathscr{Y}_{t})=\underbrace{\mathrm{var}(U_{t+h}\mid\mathscr{Y}_{t})}_{=\operatorname{var}(U_{t+h})=\sigma^{2}}+\underbrace{\mathrm{var}(\theta_{1}U_{t+h-1}\mid\mathscr{Y}_{t})}_{=\theta_{1}^{2}\operatorname{var}(U_{t+h-1})=\theta_{1}^{2}\sigma^{2}}=(1+\theta_{1}^{2})\sigma^{2}$

- **95% PI:** $\;\;$Since $Y_{t+h}\mid\mathscr{Y}_{t}\sim\mathrm{N}(0,(1+\theta_{1}^{2})\sigma^{2})\;\;\implies\;\; 0\pm1.96\times \sqrt{(1+\hat{\theta}_{1}^{2})\hat{\sigma}^{2}}$



## Exercise 3 {auto-animate="true"}
<blue>**What's going on?**</blue>
```{r echo=FALSE}
# Average interval lengths:
AvgLength <- round(apply(Length, 2, mean),2)
names(AvgLength) <- paste0("h=",1:hmax)
print(AvgLength)
```
<div class="horizontal-line"></div>

:::{.callout-tip}
<green>Consider the MA(1) model:</green> for any $t$ we have
$$Y_{t}=U_{t}+\theta_{1}U_{t-1}\quad\text{where}\quad U_{t}\sim\mathrm{N}(0,\sigma^{2})$$
To compute the 95% PI for $Y_{t+h}$, we need to know the distribution for $Y_{t+h}\mid\mathscr{Y}_{t}.$
:::
<div style="height: 0.2em;"></div>

<blue>When $h=1$, the 95% prediction interval is $\hat{\theta}_{1}u_{t}\pm 1.96\hat{\sigma}$.</blue>

<blue>When $h\ge 2$, the 95% prediction interval is $0\pm 1.96\times\sqrt{(1+\hat{\theta}_{1}^{2})\hat{\sigma}^{2}}$.</blue> 

<div class="horizontal-line"></div>

[<teal>Applying an MA(1) model produces prediction intervals of equal length for every $h>1$.</teal>]{.fragment} [The fact that the MA(1) model is misspecified in this case does not change its formula for calculation.]{.fragment}


## Exercise 4(a) {auto-animate="true"}

:::{.callout-question}
Next we well consider the properties of prediction intervals when the normality assumption is not correct. The normal and bootstrap prediction intervals will be compared.

(a) The first distribution for $U_t$ is the $\chi_{1}^{2}$ (i.e. chi-squared with 1 degree of freedom) centered to have mean 0.
:::

:::{.panel-tabset}

## True DGP
- We know that when $X\sim\chi_{k}^{2}$, $\mathbb{E}(X)=k$ and $\operatorname{var}(X)=2k$.
- To standardise it (mean 0 and unit variance), we can take the "z-score":
$$U=\frac{X-\mathbb{E}(X)}{\sqrt{\operatorname{var}(X)}}=\frac{X-1}{\sqrt{2}}$$

- The true DGP is:

$$Y_{t}=\phi_{1}Y_{t-1}+U_{t}\quad\textsf{where}\quad \color{blue}{\underbrace{\;\;U_{t}=\frac{X-1}{\sqrt{2}}\quad\textsf{and}\quad X\sim\chi_{1}^{2}\;\;}_{\chi^{2}_{1}\textsf{ centered to have mean 0}}}$$

## PDF Shape
```{r}
x <- seq(-1, 5, length.out = 500)  # support starts at -1
y <- dchisq(x + 1, df = 1)         # density of χ²(1), shifted

plot(x, y, type = "l", col = "red", lwd = 2,
     main = "", ylab = "", xlab = "", ylim = c(0, 1.5))
abline(h = 0, lty = 2)  # horizontal line at 0
```

## Full Code
The following code took me around 27 minutes and 10 seconds to run <grayed>(on a 2025 MacBook Pro, 5 minutes slower than David's "unremarkable 2011 MacBook Pro")</grayed>.

<div style="height: 0.2em;"></div>
```{r eval=FALSE}
library(forecast)

reps <- 1000
n <- 200
h <- 12
arv <- c(0.3,0.6,0.9)

Coverage <- matrix(nrow=length(arv), ncol=h)
colnames(Coverage) <- paste0("h=",1:h)
rownames(Coverage) <- paste0("ar=",arv)
CoverageBS <- Coverage
AvgLength <- Coverage
AvgLengthBS <- Coverage

et<- system.time(
for (ac in 1:length(arv)){
  ar <- arv[ac]

  Included <- matrix(nrow=reps, ncol=h)
  IncludedBS <- matrix(nrow=reps, ncol=h)
  Length <- matrix(nrow=reps, ncol=h)
  LengthBS <- matrix(nrow=reps, ncol=h)

  set.seed(3400)
  for (r in 1:reps){

    # Simulate time series
    U <- ((rnorm(n+h))^2-1)/sqrt(2)
    Y <- arima.sim(n=n+h, model=list(ar=ar), innov=U)
    Yf <- Y[(n+1):(n+h)]
    Y <- Y[1:n]

    # AR(1) estimation
    eq <- Arima(Y, order=c(1,0,0))

    # Normal prediction interval
    eqf <- forecast(eq, h=h)
    Lower <- eqf$lower[,"95%"]
    Upper <- eqf$upper[,"95%"]
    Included[r,] <- 1*(Lower<Yf & Upper>Yf)
    Length[r,] <- Upper-Lower

    # Bootstrap prediction interval
    eqf <- forecast(eq, h=h, bootstrap=TRUE)
    LowerBS <- eqf$lower[,"95%"]
    UpperBS <- eqf$upper[,"95%"]
    IncludedBS[r,] <- 1*(LowerBS<Yf & UpperBS>Yf)
    LengthBS[r,] <- UpperBS-LowerBS

  }
  Coverage[ac,] <- apply(Included, 2, mean)*100
  CoverageBS[ac,] <- apply(IncludedBS, 2, mean)*100
  AvgLength[ac,] <- round(apply(Length, 2, mean),2)
  AvgLengthBS[ac,] <- round(apply(LengthBS, 2, mean),2)
})
```
<div style="height: 1em;"></div>


## Result - Coverage
```{r echo=FALSE}
# Coverage
Coverage <- matrix(
  c(
    95,94.4,92.8,93.4,94.9,96.0,94.9,95.0,93.3,94.0,94.2,94.7,
    94,94.4,94.4,94.2,94.2,94.3,94.3,94.6,95.4,94.3,96.0,95.1,
    94,93.9,93.2,94.3,93.0,94.8,93.8,94.4,93.7,94.4,93.2,93.6
  ),
  nrow = 3, byrow = TRUE,
  dimnames = list(
    c("ar=0.3", "ar=0.6", "ar=0.9"),
    paste0("h=", 1:12)
  )
)

# CoverageBS
CoverageBS <- matrix(
  c(
    94.5,93.3,93.8,93.4,95.5,95.3,94.7,93.4,93.0,94.4,94.5,94.0,
    94.3,93.1,93.9,93.9,92.6,92.5,93.2,93.6,94.6,94.0,93.7,92.4,
    93.5,92.5,92.2,92.9,92.0,92.9,91.8,92.1,91.3,93.0,92.6,93.4
  ),
  nrow = 3, byrow = TRUE,
  dimnames = list(
    c("ar=0.3", "ar=0.6", "ar=0.9"),
    paste0("h=", 1:12)
  )
)
```
```{r}
print(Coverage)     # Normal Assumption (Misspecified) 
print(CoverageBS)   # Bootstrapped 
```
<div class="horizontal-line"></div>
- The coverage results are very similar across the boards.
  - <blue>Note:</blue> Remember these are just simulation estimates from only 1000 replications, and don't over-interpret small differences.
- Nevertheless there might be a slight deficiency of about 1% in  the coverage of the bootstrap intervals relative to the normal intervals when $\phi=0.9$.
- Otherwise, there is little to choose between the two intervals on this criterion.
  - This is quite interesting because theoretically the bootstrap intervals are valid in this case but the normal intervals are not.
  - <blue>Note:</blue> CLT does not provide an approximation for non-normal data when computing prediction intervals. 

## Result - Lengths
```{r echo=FALSE}
# AvgLength
AvgLength <- matrix(
  c(
    3.90,4.06,4.08,4.08,4.08,4.08,4.08,4.08,4.08,4.08,4.08,4.08,
    3.91,4.54,4.75,4.82,4.85,4.86,4.86,4.87,4.87,4.87,4.87,4.87,
    3.91,5.22,6.05,6.64,7.07,7.39,7.64,7.83,7.98,8.10,8.20,8.27
  ),
  nrow = 3, byrow = TRUE,
  dimnames = list(
    c("ar=0.3", "ar=0.6", "ar=0.9"),
    paste0("h=", 1:12)
  )
)

# AvgLengthBS
AvgLengthBS <- matrix(
  c(
    3.61,3.81,3.86,3.86,3.87,3.85,3.86,3.87,3.86,3.86,3.86,3.86,
    3.62,4.28,4.53,4.63,4.66,4.68,4.69,4.70,4.69,4.69,4.70,4.70,
    3.62,4.93,5.80,6.41,6.86,7.19,7.45,7.65,7.82,7.94,8.04,8.13
  ),
  nrow = 3, byrow = TRUE,
  dimnames = list(
    c("ar=0.3", "ar=0.6", "ar=0.9"),
    paste0("h=", 1:12)
  )
)
```
```{r}
print(AvgLength)    # Normal Assumption (Misspecified)
print(AvgLengthBS)  # Bootstrapped 
```
<div class="horizontal-line"></div>

The bootstrap intervals can be as much as 10% shorter for small $h$, although this advantage reduces to 5% or less as $h$ increases.

:::


## Exercise 4(a) {auto-animate="true"}

:::{.callout-question}
Next we well consider the properties of prediction intervals when the normality assumption is not correct. The normal and bootstrap prediction intervals will be compared.

(a) The first distribution for $U_t$ is the $\chi_{1}^{2}$ (i.e. chi-squared with 1 degree of freedom) centered to have mean 0.
:::
<div style="height: 0.5em;"></div>
<blue>**Overall:**</blue>

:::{.incremental}
- There are not major differences between the two intervals and the choice would be subjective.
- The advantage in length may suggest some *weak preference* for the bootstrap intervals in this particular model.
:::


## Exercise 4(b) {auto-animate="true"}

:::{.callout-question}
Next we well consider the properties of prediction intervals when the normality assumption is not correct. The normal and bootstrap prediction intervals will be compared.

(b) The second distribution for $U_t$ is the $t$ distribution with 4 degrees of freedom.
:::

:::{.panel-tabset}

## True DGP

The true DGP is:
$$Y_{t}=\phi_{1}Y_{t-1}+U_{t}\quad\textsf{where}\quad \color{blue}{U_{t}\sim t_{4}}$$
This distribution has fatter tails (higher kurtosis) than $\mathrm{N}(0,1)$:
```{r}
x <- seq(-4, 4, length.out = 1000)
y_t <- dt(x, df = 4)
y_norm <- dnorm(x)

plot(x, y_t, type = "l", col = "red", lwd = 2,
     ylim = c(0, 0.45),
     ylab = "", xlab = "", main = "")
lines(x, y_norm, col = "blue", lty = 2, lwd = 2)
abline(h = 0, lty = 3)
legend("topright", legend = c("t (df=4)", "Normal(0,1)"), 
       col = c("red", "blue"), lty = c(1, 2), lwd = 2)
```
## Full Code

```{r eval=FALSE}
library(forecast)

reps <- 1000
n <- 200
h <- 12
arv <- c(0.3,0.6,0.9)

Coverage <- matrix(nrow=length(arv), ncol=h)
colnames(Coverage) <- paste0("h=",1:h)
rownames(Coverage) <- paste0("ar=",arv)
CoverageBS <- Coverage
AvgLength <- Coverage
AvgLengthBS <- Coverage

et<- system.time(
for (ac in 1:length(arv)){
  ar <- arv[ac]

  Included <- matrix(nrow=reps, ncol=h)
  IncludedBS <- matrix(nrow=reps, ncol=h)
  Length <- matrix(nrow=reps, ncol=h)
  LengthBS <- matrix(nrow=reps, ncol=h)

  set.seed(3400)
  for (r in 1:reps){

    # Simulate time series
    U <- rt(n=n+h, df=4)
    Y <- arima.sim(n=n+h, model=list(ar=ar), innov=U)
    Yf <- Y[(n+1):(n+h)]
    Y <- Y[1:n]

    # AR(1) estimation
    eq <- Arima(Y, order=c(1,0,0))

    # Normal prediction interval
    eqf <- forecast(eq, h=h)
    Lower <- eqf$lower[,"95%"]
    Upper <- eqf$upper[,"95%"]
    Included[r,] <- 1*(Lower<Yf & Upper>Yf)
    Length[r,] <- Upper-Lower

    # Bootstrap prediction interval
    eqf <- forecast(eq, h=h, bootstrap=TRUE)
    LowerBS <- eqf$lower[,"95%"]
    UpperBS <- eqf$upper[,"95%"]
    IncludedBS[r,] <- 1*(LowerBS<Yf & UpperBS>Yf)
    LengthBS[r,] <- UpperBS-LowerBS

  }
  Coverage[ac,] <- apply(Included, 2, mean)*100
  CoverageBS[ac,] <- apply(IncludedBS, 2, mean)*100
  AvgLength[ac,] <- round(apply(Length, 2, mean),2)
  AvgLengthBS[ac,] <- round(apply(LengthBS, 2, mean),2)
})
```
<div style="height: 1em;"></div>

```{r echo=FALSE}
# Normal PI coverage
Coverage_normal <- matrix(
  c(
    94.9,95.0,94.8,94.0,94.6,94.8,93.3,94.0,94.5,94.4,95.9,94.2,
    95.2,93.8,94.2,93.6,92.1,94.5,95.2,94.0,94.0,94.9,93.6,93.5,
    94.6,95.4,93.7,93.4,93.4,93.4,93.4,91.3,91.5,90.8,91.0,90.8
  ),
  nrow = 3, byrow = TRUE
)
colnames(Coverage_normal) <- paste0("h=",1:12)
rownames(Coverage_normal) <- c("ar=0.3","ar=0.6","ar=0.9")
Coverage <- Coverage_normal

# Bootstrap PI coverage
Coverage_bootstrap <- matrix(
  c(
    94.8,94.6,95.0,94.1,94.3,94.8,93.5,94.2,94.3,94.5,95.4,93.6,
    95.7,94.3,93.5,93.4,92.5,94.4,95.3,94.2,93.7,94.6,93.3,93.9,
    94.2,95.4,93.8,93.7,92.9,93.1,93.0,91.1,91.3,90.2,90.4,90.1
  ),
  nrow = 3, byrow = TRUE
)
colnames(Coverage_bootstrap) <- paste0("h=",1:12)
rownames(Coverage_bootstrap) <- c("ar=0.3","ar=0.6","ar=0.9")
CoverageBS <- Coverage_bootstrap

# Normal PI average length
AvgLength_normal <- matrix(
  c(
    5.48,5.71,5.73,5.74,5.74,5.74,5.74,5.74,5.74,5.74,5.74,5.74,
    5.45,6.32,6.60,6.70,6.74,6.76,6.76,6.76,6.77,6.77,6.77,6.77,
    5.52,7.35,8.50,9.31,9.89,10.33,10.66,10.92,11.12,11.28,11.40,11.50
  ),
  nrow = 3, byrow = TRUE
)
colnames(AvgLength_normal) <- paste0("h=",1:12)
rownames(AvgLength_normal) <- c("ar=0.3","ar=0.6","ar=0.9")
AvgLength <- AvgLength_normal

# Bootstrap PI average length
AvgLength_bootstrap <- matrix(
  c(
    5.54,5.77,5.80,5.80,5.79,5.81,5.79,5.80,5.80,5.80,5.80,5.80,
    5.56,6.43,6.72,6.81,6.85,6.86,6.87,6.87,6.86,6.85,6.86,6.87,
    5.55,7.44,8.64,9.46,10.06,10.56,10.90,11.13,11.33,11.47,11.59,11.69
  ),
  nrow = 3, byrow = TRUE
)
colnames(AvgLength_bootstrap) <- paste0("h=",1:12)
rownames(AvgLength_bootstrap) <- c("ar=0.3","ar=0.6","ar=0.9")
AvgLengthBS <- AvgLength_bootstrap
```

## Result - Coverage
```{r}
print(Coverage)     # Normal Assumption (Misspecified) 
print(CoverageBS)   # Bootstrapped 
```
<div class="horizontal-line"></div>

<blue>**Comments:**</blue>

- One finding to note is that the coverage properties deteriorate for larger $h$ when $\phi_{1}=0.9$.

- The finite sample statistical properties of prediction with strong autocorrelation (i.e., large $\phi_{1}$) can require larger samples for the asymptotic properties (i.e. 95% coverage) to apply.
<div style="height: 0.5em;"></div>

## Result - Lengths
```{r}
print(AvgLength)    # Normal Assumption (Misspecified)
print(AvgLengthBS)  # Bootstrapped 
```
<div class="horizontal-line"></div>

The average lengths of the prediction intervals increases with $h$ and $\phi_{1}$ <grayed>(the same applied in part a)</grayed> <teal>$\implies$ Long-horizon prediction is more difficult when time series are more strongly autocorrelated.</teal>
<div style="height: 0.5em;"></div>

## Conclusion

- Overall, there is no advantage to be found for the bootstrap intervals in this model. If anything they are slightly longer.

- Despite the non-normality, the normal prediction intervals work quite well in this case.

- This could be due to the shape of non-normality is different in this case, as the distribution is symmetric.

:::



## {data-background-color="#000F46"}
<div style="height: 85vh; display: flex; align-items: center; justify-content: center; color: white; font-size: 2em;">
Thank you for coming!
</div>



<script>
document.querySelectorAll("myimage").forEach(el => {
  const src = el.getAttribute("src");
  const width = el.getAttribute("width") || "100%";

  const img = document.createElement("img");
  img.src = src;
  img.style.width = width;
  img.style.maxWidth = "100%";  // prevent overflow
  el.style.display = "block";
  el.style.textAlign = "center";
  
  el.appendChild(img);
});
</script>